Requirements Specification

### 2025-05-13 nbs/development/architecture.qmd
docs(development): update parameter recovery validation requirements
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@

@@

1. **PyroVelocityModel**: Main model class that composes specialized component models
2. **Dynamics Models**: Implement the mathematical models of RNA dynamics
   - StandardDynamicsModel: Standard implementation of RNA dynamics
   - LegacyDynamicsModel: Compatibility implementation matching the legacy model
3. **Prior Models**: Define prior distributions for model parameters
   - LogNormalPriorModel: Uses log-normal distributions for parameters
4. **Likelihood Models**: Define observation distributions
   - PoissonLikelihoodModel: Uses Poisson distributions for RNA counts
   - LegacyLikelihoodModel: Compatibility implementation matching the legacy model
5. **Observation Models**: Handle data preprocessing and transformation
   - StandardObservationModel: Standard implementation for observations
6. **Inference Guides**: Create guide functions for variational inference
   - AutoGuideFactory: Creates guides for the modular implementation
   - LegacyAutoGuideFactory: Compatibility implementation matching the legacy guide
7. **Registry System**: Enables component registration and retrieval
8. **Factory System**: Creates components and models from configurations

@@

| Component | Major Function | Description | Related Core Function |
|-----------|----------------|-------------|----------------------|
| PyroVelocityModel | forward | Runs the model forward to compute expected RNA counts | FN-03 |
| PyroVelocityModel | train | Trains the model using variational inference | FN-03 |
| PyroVelocityModel | generate_posterior_samples | Generates posterior samples for uncertainty quantification | FN-04 |
| PyroVelocityModel | compute_velocity | Computes RNA velocity from posterior samples | FN-05 |
| PyroVelocityModel | store_results_in_anndata | Stores results in AnnData objects | FN-08 |
| DynamicsModel | forward | Computes expected unspliced and spliced RNA counts | FN-05 |
| PriorModel | forward | Samples model parameters from prior distributions | FN-03 |
| LikelihoodModel | forward | Defines likelihood distributions for observed data | FN-03 |
| ObservationModel | forward | Transforms observed data for model input | FN-01 |
| InferenceGuide | create_guide | Creates guide functions for variational inference | FN-03 |

@@

| Decision ID | Decision | Rationale | Alternatives Considered | Impact |
|-------------|----------|-----------|-------------------------|--------|
| AD-01 | Protocol-First Architecture | Enables composition over inheritance, maximum flexibility, and alignment with functional programming principles | Abstract Base Classes | Complete separation between interface and implementation, maximum flexibility, simplified testing |
| AD-02 | Context-Based Component Communication | Flexible data sharing without tight coupling, clear data flow | Direct method parameters | Easy to extend with new data, potential for inconsistent context structure |
| AD-03 | Immutable State Pattern | Thread safety, predictable behavior, functional programming style | Mutable state | Simplified reasoning about code, reduced side effects |
| AD-04 | Registry Pattern | Dynamic component discovery, extensibility | Direct imports | Decoupling between component implementations and usage |
| AD-05 | Factory Pattern | Encapsulation of creation logic, configuration management | Direct instantiation | Simplified client code, centralized creation logic |

@@

@@

>>>>

```{mermaid}
graph TB
    Main["PyroVelocityModel"] -->|contains| Dynamics["Dynamics Models"]
    Main -->|contains| Priors["Prior Models"]
    Main -->|contains| Likelihoods["Likelihood Models"]
    Main -->|contains| Observations["Observation Models"]
    Main -->|contains| Guides["Inference Guides"]
    Main -->|contains| Registry["Registry System"]
    Main -->|contains| Factory["Factory System"]
    Main -->|uses| Validation["Validation Framework"]

@@

    Validation -->|contains| ParameterRecovery["Parameter Recovery"]
    Validation -->|contains| ModelComparison["Model Comparison"]
    Validation -->|contains| ValidationMetrics["Validation Metrics"]

@@

    class Main main;
    class Dynamics,Priors,Likelihoods,Observations,Guides component;
    class DynamicsInterface,PriorsInterface,LikelihoodsInterface,ObservationsInterface,GuidesInterface interface;
    class Registry,Factory system;
    class Validation,ParameterRecovery,ModelComparison,ValidationMetrics validation;
```

### Core Components {#sec-overview-core-components}

1. **PyroVelocityModel**: Main model class that composes specialized component models
2. **Dynamics Models**: Implement the mathematical models of RNA dynamics
   - StandardDynamicsModel: Standard implementation of RNA dynamics
   - LegacyDynamicsModel: Compatibility implementation matching the legacy model
3. **Prior Models**: Define prior distributions for model parameters
   - LogNormalPriorModel: Uses log-normal distributions for parameters
4. **Likelihood Models**: Define observation distributions
   - PoissonLikelihoodModel: Uses Poisson distributions for RNA counts
   - LegacyLikelihoodModel: Compatibility implementation matching the legacy model
5. **Observation Models**: Handle data preprocessing and transformation
   - StandardObservationModel: Standard implementation for observations
6. **Inference Guides**: Create guide functions for variational inference
   - AutoGuideFactory: Creates guides for the modular implementation
   - LegacyAutoGuideFactory: Compatibility implementation matching the legacy guide
7. **Registry System**: Enables component registration and retrieval
8. **Factory System**: Creates components and models from configurations
9. **Validation Framework**: Provides tools for model validation and comparison
   - Parameter Recovery: Validates model's ability to recover known parameters from synthetic data
   - Model Comparison: Compares different model implementations and configurations
   - Validation Metrics: Computes metrics for quantifying validation performance

@@

| Component | Major Function | Description | Related Core Function |
|-----------|----------------|-------------|----------------------|
| PyroVelocityModel | forward | Runs the model forward to compute expected RNA counts | FN-03 |
| PyroVelocityModel | train | Trains the model using variational inference | FN-03 |
| PyroVelocityModel | generate_posterior_samples | Generates posterior samples for uncertainty quantification | FN-04 |
| PyroVelocityModel | compute_velocity | Computes RNA velocity from posterior samples | FN-05 |
| PyroVelocityModel | store_results_in_anndata | Stores results in AnnData objects | FN-08 |
| DynamicsModel | forward | Computes expected unspliced and spliced RNA counts | FN-05 |
| PriorModel | forward | Samples model parameters from prior distributions | FN-03 |
| LikelihoodModel | forward | Defines likelihood distributions for observed data | FN-03 |
| ObservationModel | forward | Transforms observed data for model input | FN-01 |
| InferenceGuide | create_guide | Creates guide functions for variational inference | FN-03 |
| ParameterRecovery | generate_synthetic_data | Generates synthetic data with known parameters | FN-03 |
| ParameterRecovery | validate_parameter_recovery | Validates model's ability to recover known parameters | FN-03 |
| ModelComparison | compare_parameters | Compares parameter estimates between implementations | FN-04 |
| ValidationMetrics | compute_parameter_metrics | Computes metrics for parameter recovery validation | FN-04 |

@@

| Decision ID | Decision | Rationale | Alternatives Considered | Impact |
|-------------|----------|-----------|-------------------------|--------|
| AD-01 | Protocol-First Architecture | Enables composition over inheritance, maximum flexibility, and alignment with functional programming principles | Abstract Base Classes | Complete separation between interface and implementation, maximum flexibility, simplified testing |
| AD-02 | Context-Based Component Communication | Flexible data sharing without tight coupling, clear data flow | Direct method parameters | Easy to extend with new data, potential for inconsistent context structure |
| AD-03 | Immutable State Pattern | Thread safety, predictable behavior, functional programming style | Mutable state | Simplified reasoning about code, reduced side effects |
| AD-04 | Registry Pattern | Dynamic component discovery, extensibility | Direct imports | Decoupling between component implementations and usage |
| AD-05 | Factory Pattern | Encapsulation of creation logic, configuration management | Direct instantiation | Simplified client code, centralized creation logic |
| AD-06 | Parameter Recovery Validation | Validate model's ability to recover known parameters from synthetic data before applying to real data | Direct comparison to legacy implementation only | Ensures model is internally consistent and capable of recovering parameters, prerequisite for robust model comparison |

@@

**Method ID:** CM-02

**Method Name:** Parameter Recovery Validation

**Related Function:** FN-03 (Parameter Recovery Validation)

**Description:** Validates model's ability to recover known parameters from synthetic data

- Parameter generation: Samples parameter sets from prior distributions
- Synthetic data generation: Creates synthetic data using known parameters
- Inference: Performs inference on synthetic data to estimate parameters
- Evaluation: Compares estimated parameters to original known values
- Metrics computation: Calculates recovery accuracy, precision, and coverage metrics

- Libraries/frameworks used: PyTorch, Pyro, NumPy, SciPy
- Computational complexity: O(num_parameter_sets * num_samples * max_sample_size * num_genes)
- Memory requirements: O(num_parameter_sets * num_samples * max_sample_size * num_genes)
- Parallelization strategy: Parallel processing across parameter sets and samples

@@

**Interface ID:** CIF-02

**Interface Name:** ParameterRecoveryValidator Protocol

**Description:** Interface for parameter recovery validation

```python
@runtime_checkable
class ParameterRecoveryValidator(Protocol):
    def generate_synthetic_data(
        self,
        model_config: Dict[str, Any],
        prior_config: Dict[str, Any],
        num_cells: int,
        num_genes: int,
        **kwargs
    ) -> Dict[str, Any]:
        """
        Generate synthetic data with known parameters.
        """
        ...

    def validate_parameter_recovery(
        self,
        model_config: Dict[str, Any],
        synthetic_data: Dict[str, Any],
        **kwargs
    ) -> Dict[str, Any]:
        """
        Validate model's ability to recover known parameters.
        """
        ...
```

| Parameter | Type | Description | Default Value | Constraints |
|-----------|------|-------------|---------------|-------------|
| model_config | Dict[str, Any] | Model configuration | None | Must specify model components |
| prior_config | Dict[str, Any] | Prior distribution configuration | None | Must specify prior distributions for all parameters |
| num_cells | int | Number of cells in synthetic data | None | > 0 |
| num_genes | int | Number of genes in synthetic data | None | > 0 |
| synthetic_data | Dict[str, Any] | Synthetic data with known parameters | None | Must contain u_obs, s_obs, true_parameters |

| Return Value | Type | Description | Constraints |
|--------------|------|-------------|-------------|
| synthetic_data | Dict[str, Any] | Synthetic data with known parameters | Must contain u_obs, s_obs, true_parameters |
| validation_results | Dict[str, Any] | Validation metrics and results | Must contain recovery_metrics, coverage_metrics |

- Missing parameters: Raises ValueError with descriptive message
- Invalid parameters: Validates parameters and raises ValueError for invalid values
- Inference failure: Provides diagnostic information for inference failures

### 2025-05-13 nbs/development/context.qmd
docs(development): update parameter recovery validation requirements
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

>>>>

### Parameter Recovery Validation Workflow

- **Purpose**: Validate model's ability to recover known parameters from synthetic data
- **Inputs**: Model specification and prior distributions
- **Steps**:
  1. Generate known parameter sets from prior distributions
  2. Create synthetic data using these parameters and the model's generative process
  3. Perform inference on synthetic data to estimate parameters
  4. Compare estimated parameters to original known values
  5. Analyze recovery accuracy, precision, and coverage
- **Outputs**: Validation metrics and visualizations demonstrating parameter recovery performance

### 2025-05-13 nbs/development/requirements.qmd
docs(development): update parameter recovery validation requirements
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

| Capability ID | Capability Name | Description | Priority |
|--------------|-----------------|-------------|----------|
| CAP-01 | Probabilistic RNA Velocity | Bayesian modeling of RNA velocity with uncertainty quantification | High |
| CAP-02 | Parameter Inference | Inference of biologically meaningful parameters from scRNA-seq data | High |
| CAP-03 | Uncertainty Visualization | Visualization of uncertainty in velocity estimates and trajectories | High |
| CAP-04 | AnnData Integration | Seamless integration with AnnData objects and the scanpy ecosystem | High |
| CAP-05 | Modular Architecture | Flexible, component-based architecture with interchangeable parts | High |
| CAP-06 | JAX/NumPyro Backend | High-performance implementation using JAX/NumPyro | Medium |
| CAP-07 | Workflow Integration | Integration with workflow systems for reproducible analysis | Medium |

@@

- **Data Preprocessing**: Scientists prepare single-cell data for velocity analysis
- **Model Configuration**: Scientists configure model components and parameters
- **Model Training**: Scientists train the model using variational inference
- **Posterior Analysis**: Scientists analyze posterior distributions of parameters
- **Velocity Computation**: Scientists compute RNA velocity with uncertainty
- **Result Visualization**: Scientists visualize velocity and uncertainty
- **Integration with Workflows**: Scientists integrate velocity analysis into larger pipelines

```{mermaid}
graph TD
    A[Data Preprocessing] --> B[Model Configuration]
    B --> C[Model Training]
    C --> D[Posterior Analysis]
    D --> E[Velocity Computation]
    E --> F[Result Visualization]
    F --> G[Integration with Workflows]

    class A,B,C,D,E,F,G scenario;
```

@@

| Scenario ID | Scenario Name | Primary Actor | Description | Pre-conditions | Post-conditions |
|-------------|---------------|---------------|-------------|----------------|-----------------|
| US-01 | Data Preprocessing | A-01, A-02 | Prepare single-cell data for velocity analysis | Raw count data available | AnnData object with spliced/unspliced counts |
| US-02 | Model Configuration | A-01, A-04 | Configure model components and parameters | Preprocessed data available | Configured model ready for training |
| US-03 | Model Training | A-01, A-02 | Train the model using variational inference | Configured model and data available | Trained model with optimized parameters |
| US-04 | Posterior Analysis | A-01, A-04 | Analyze posterior distributions of parameters | Trained model available | Parameter distributions and diagnostics |
| US-05 | Velocity Computation | A-01, A-02 | Compute RNA velocity with uncertainty | Posterior samples available | Velocity estimates with uncertainty |
| US-06 | Result Visualization | A-01, A-03 | Visualize velocity and uncertainty | Velocity estimates available | Visualizations for interpretation |
| US-07 | Integration with Workflows | A-02 | Integrate velocity analysis into larger pipelines | Trained model and analysis code available | Integrated workflow |

@@

**Usage Scenario:** US-03 (Model Training)

@@

**Workflow Name:** RNA Velocity Analysis Workflow

**Description:** Complete workflow for RNA velocity analysis from raw data to interpretation

**Participants:** A-01, A-02, A-03

@@

1. Data Preprocessing: US-01
2. Model Configuration: US-02
3. Model Training: US-03
4. Posterior Analysis: US-04
5. Velocity Computation: US-05
6. Result Visualization: US-06

@@

* For large datasets, users may subsample genes or cells before model training
* For complex biological systems, users may use nonlinear dynamics models
* For integration with other analyses, users may export velocity to other tools

@@

@@

@@

| Data Object ID | Data Object Name | Description | Related Scientific Data Object | Persistence |
|----------------|------------------|-------------|-------------------------------|------------|
| DO-01 | AnnData | Annotated data matrix for single-cell data | AnnData | Yes |
| DO-02 | CountMatrix | Matrix of RNA counts | CountMatrix | Yes |
| DO-03 | ModelParameters | Parameters of the RNA velocity model | ModelParameters | Yes |
| DO-04 | PosteriorSamples | Samples from parameter posterior distributions | PosteriorSamples | Yes |
| DO-05 | VelocityEstimates | RNA velocity vectors for each cell | VelocityEstimates | Yes |
| DO-06 | ModelState | Immutable state container for model components | N/A | Yes |
| DO-07 | ComponentConfig | Configuration for model components | N/A | Yes |

@@

@@

| Relationship ID | From Object | To Object | Type | Cardinality | Description |
|-----------------|-------------|-----------|------|-------------|-------------|
| REL-01 | AnnData | CountMatrix | Composition | 1:N | AnnData contains multiple count matrices (X, layers) |
| REL-02 | PyroVelocityModel | ModelParameters | Association | 1:1 | Model uses parameters for computation |
| REL-03 | PyroVelocityModel | PosteriorSamples | Association | 1:1 | Model generates posterior samples |
| REL-04 | PosteriorSamples | VelocityEstimates | Derivation | 1:1 | Velocity is computed from posterior samples |
| REL-05 | AnnData | VelocityEstimates | Association | 1:1 | Velocity estimates are stored in AnnData |
| REL-06 | PyroVelocityModel | ModelState | Composition | 1:1 | Model contains state |
| REL-07 | ModelState | ComponentConfig | Composition | 1:N | State contains component configurations |

```{mermaid}
classDiagram
    AnnData *-- CountMatrix : contains
    PyroVelocityModel o-- ModelParameters : uses
    PyroVelocityModel --> PosteriorSamples : generates
    PosteriorSamples --> VelocityEstimates : computes
    AnnData o-- VelocityEstimates : stores
    PyroVelocityModel *-- ModelState : contains
    ModelState *-- ComponentConfig : contains

@@

    class ComponentConfig {
        +str name
        +Dict params
    }
```

>>>>

| Capability ID | Capability Name | Description | Priority |
|--------------|-----------------|-------------|----------|
| CAP-01 | Probabilistic RNA Velocity | Bayesian modeling of RNA velocity with uncertainty quantification | High |
| CAP-02 | Parameter Inference | Inference of biologically meaningful parameters from scRNA-seq data | High |
| CAP-03 | Uncertainty Visualization | Visualization of uncertainty in velocity estimates and trajectories | High |
| CAP-04 | AnnData Integration | Seamless integration with AnnData objects and the scanpy ecosystem | High |
| CAP-05 | Modular Architecture | Flexible, component-based architecture with interchangeable parts | High |
| CAP-06 | Parameter Recovery Validation | Validation of model's ability to recover known parameters from synthetic data | High |
| CAP-07 | Model Comparison | Comparison of different model implementations and configurations | High |
| CAP-08 | JAX/NumPyro Backend | High-performance implementation using JAX/NumPyro | Medium |
| CAP-09 | Workflow Integration | Integration with workflow systems for reproducible analysis | Medium |

@@

- **Data Preprocessing**: Scientists prepare single-cell data for velocity analysis
- **Model Configuration**: Scientists configure model components and parameters
- **Parameter Recovery Validation**: Scientists validate model's ability to recover known parameters
- **Model Training**: Scientists train the model using variational inference
- **Posterior Analysis**: Scientists analyze posterior distributions of parameters
- **Velocity Computation**: Scientists compute RNA velocity with uncertainty
- **Result Visualization**: Scientists visualize velocity and uncertainty
- **Model Comparison**: Scientists compare different model implementations and configurations
- **Integration with Workflows**: Scientists integrate velocity analysis into larger pipelines

```{mermaid}
graph TD
    A[Data Preprocessing] --> B[Model Configuration]
    B --> V[Parameter Recovery Validation]
    V --> C[Model Training]
    C --> D[Posterior Analysis]
    D --> E[Velocity Computation]
    E --> F[Result Visualization]
    F --> M[Model Comparison]
    M --> G[Integration with Workflows]

    class A,B,V,C,D,E,F,M,G scenario;
```

@@

| Scenario ID | Scenario Name | Primary Actor | Description | Pre-conditions | Post-conditions |
|-------------|---------------|---------------|-------------|----------------|-----------------|
| US-01 | Data Preprocessing | A-01, A-02 | Prepare single-cell data for velocity analysis | Raw count data available | AnnData object with spliced/unspliced counts |
| US-02 | Model Configuration | A-01, A-04 | Configure model components and parameters | Preprocessed data available | Configured model ready for training |
| US-03 | Parameter Recovery Validation | A-01, A-04 | Validate model's ability to recover known parameters from synthetic data | Model specification and prior distributions available | Validation metrics and visualizations demonstrating parameter recovery performance |
| US-04 | Model Training | A-01, A-02 | Train the model using variational inference | Configured model and data available | Trained model with optimized parameters |
| US-05 | Posterior Analysis | A-01, A-04 | Analyze posterior distributions of parameters | Trained model available | Parameter distributions and diagnostics |
| US-06 | Velocity Computation | A-01, A-02 | Compute RNA velocity with uncertainty | Posterior samples available | Velocity estimates with uncertainty |
| US-07 | Result Visualization | A-01, A-03 | Visualize velocity and uncertainty | Velocity estimates available | Visualizations for interpretation |
| US-08 | Model Comparison | A-01, A-04 | Compare different model implementations and configurations | Multiple trained models available | Comparison metrics and visualizations |
| US-09 | Integration with Workflows | A-02 | Integrate velocity analysis into larger pipelines | Trained model and analysis code available | Integrated workflow |

@@

**Usage Scenario:** US-03 (Parameter Recovery Validation)

**Description:** Validating a PyroVelocity model's ability to recover known parameters from synthetic data

**Triggering Event:** User initiates parameter recovery validation

1. User specifies model configuration and prior distributions
2. System generates known parameter sets from prior distributions
3. System creates synthetic data using these parameters and the model's generative process
4. System performs inference on synthetic data to estimate parameters
5. System compares estimated parameters to original known values
6. System analyzes recovery accuracy, precision, and coverage
7. System generates validation metrics and visualizations
8. User evaluates parameter recovery performance

* At step 1, if user wants to test specific parameter values, they provide custom parameter sets
* At step 3, if user wants to test robustness, they specify noise levels to add to synthetic data

* At step 4, if inference fails to converge, system provides diagnostic information
* At step 5, if parameters are not recoverable, system identifies problematic parameters

**Scenario ID:** FS-02

**Usage Scenario:** US-04 (Model Training)

@@

**Workflow Name:** RNA Velocity Analysis Workflow with Parameter Recovery Validation

**Description:** Complete workflow for RNA velocity analysis from raw data to interpretation, including parameter recovery validation

**Participants:** A-01, A-02, A-03, A-04

@@

1. Data Preprocessing: US-01
2. Model Configuration: US-02
3. Parameter Recovery Validation: US-03
4. Model Training: US-04
5. Posterior Analysis: US-05
6. Velocity Computation: US-06
7. Result Visualization: US-07
8. Model Comparison: US-08 (optional)

@@

* For large datasets, users may subsample genes or cells before model training
* For complex biological systems, users may use nonlinear dynamics models
* For integration with other analyses, users may export velocity to other tools
* For comprehensive validation, users may test multiple parameter sets and noise levels
* For model comparison, users may compare different model implementations and configurations

**Workflow ID:** WF-02

**Workflow Name:** Parameter Recovery Validation Workflow

**Description:** Focused workflow for validating model's ability to recover known parameters from synthetic data

**Participants:** A-01, A-04

1. Model Configuration: US-02
2. Parameter Recovery Validation: US-03
3. Result Visualization: US-07
4. Model Comparison: US-08 (if comparing multiple models)

* For comprehensive validation, users may test multiple parameter sets and noise levels
* For robustness testing, users may vary sample sizes and noise levels
* For model development, users may iterate on model design based on validation results

@@

**Algorithm Name:** Parameter Recovery Validation

**Description:** Validation of model's ability to recover known parameters from synthetic data

```
Recovery Error = ||θ_true - θ_estimated||
Coverage = P(θ_true ∈ CI(θ_estimated))
```

Where:
- θ_true are the true parameter values
- θ_estimated are the estimated parameter values
- CI(θ_estimated) is the credible interval for the estimated parameters
- ||·|| is a suitable distance metric (e.g., L2 norm)

- Model: Probabilistic model specification
- Prior: Prior distributions for parameters
- Num_parameter_sets: Number of parameter sets to test
- Num_samples: Number of synthetic datasets per parameter set
- Sample_sizes: List of sample sizes to test
- Noise_levels: List of noise levels to test

- Recovery accuracy metrics (MSE, correlation, etc.)
- Recovery precision metrics (width of credible intervals)
- Coverage metrics (percentage of true parameters within credible intervals)
- Bias metrics (systematic deviation from true values)
- Visualizations of parameter recovery performance

- Time Complexity: O(num_parameter_sets * num_samples * max_sample_size * num_genes)
- Space Complexity: O(num_parameter_sets * num_samples * max_sample_size * num_genes)
- Parallelization: Embarrassingly parallel across parameter sets and samples

- Parameter recovery validation specification
- Principled Bayesian workflow literature

**Algorithm ID:** ALG-03

@@

**Description:** Interface for parameter recovery validation

| Parameter | Type | Description | Constraints |
|-----------|------|-------------|------------|
| model_config | Dict[str, Any] | Model configuration | Must specify model components |
| prior_config | Dict[str, Any] | Prior distribution configuration | Must specify prior distributions for all parameters |
| num_parameter_sets | int | Number of parameter sets to test | > 0 |
| num_samples | int | Number of synthetic datasets per parameter set | > 0 |
| sample_sizes | List[int] | List of sample sizes to test | All values > 0 |
| noise_levels | List[float] | List of noise levels to test | All values >= 0 |
| use_gpu | bool | Whether to use GPU acceleration | True/False |

| Parameter | Type | Description |
|-----------|------|-------------|
| validation_results | Dict[str, Any] | Validation metrics and results |
| parameter_recovery_plots | Dict[str, Figure] | Visualizations of parameter recovery |
| recovery_summary | Dict[str, float] | Summary statistics of recovery performance |

| Error Code | Description | Recovery Action |
|------------|-------------|----------------|
| ValueError | Invalid model or prior configuration | Correct configuration |
| RuntimeError | Inference failure | Adjust inference parameters or model complexity |
| MemoryError | Out of memory | Reduce number of parameter sets or sample sizes |

**Interface ID:** CI-03

**Algorithm:** ALG-03

@@

| Data Object ID | Data Object Name | Description | Related Scientific Data Object | Persistence |
|----------------|------------------|-------------|-------------------------------|------------|
| DO-01 | AnnData | Annotated data matrix for single-cell data | AnnData | Yes |
| DO-02 | CountMatrix | Matrix of RNA counts | CountMatrix | Yes |
| DO-03 | ModelParameters | Parameters of the RNA velocity model | ModelParameters | Yes |
| DO-04 | PosteriorSamples | Samples from parameter posterior distributions | PosteriorSamples | Yes |
| DO-05 | VelocityEstimates | RNA velocity vectors for each cell | VelocityEstimates | Yes |
| DO-06 | ModelState | Immutable state container for model components | N/A | Yes |
| DO-07 | ComponentConfig | Configuration for model components | N/A | Yes |
| DO-08 | SyntheticData | Synthetic data generated with known parameters | N/A | Yes |
| DO-09 | ParameterRecoveryResults | Results of parameter recovery validation | N/A | Yes |

@@

**Data Object:** DO-08 (SyntheticData)

| Attribute ID | Attribute Name | Description | Type | Units | Constraints | Default Value |
|--------------|----------------|-------------|------|-------|-------------|---------------|
| ATTR-15 | u_obs | Observed unspliced counts | Tensor | Counts | Non-negative | None |
| ATTR-16 | s_obs | Observed spliced counts | Tensor | Counts | Non-negative | None |
| ATTR-17 | true_parameters | True parameter values used to generate data | Dict[str, Tensor] | Various | Non-negative | None |
| ATTR-18 | sample_size | Number of cells in the synthetic dataset | int | Count | > 0 | None |
| ATTR-19 | noise_level | Noise level used in data generation | float | Dimensionless | >= 0 | 0.0 |
| ATTR-20 | seed | Random seed used for reproducibility | int | Dimensionless | Any | None |

**Data Object:** DO-09 (ParameterRecoveryResults)

| Attribute ID | Attribute Name | Description | Type | Units | Constraints | Default Value |
|--------------|----------------|-------------|------|-------|-------------|---------------|
| ATTR-21 | true_parameters | True parameter values | Dict[str, Tensor] | Various | Non-negative | None |
| ATTR-22 | estimated_parameters | Estimated parameter values | Dict[str, Tensor] | Various | Non-negative | None |
| ATTR-23 | credible_intervals | Credible intervals for estimated parameters | Dict[str, Tuple[Tensor, Tensor]] | Various | Lower < Upper | None |
| ATTR-24 | recovery_metrics | Metrics quantifying parameter recovery performance | Dict[str, Dict[str, float]] | Various | None | None |
| ATTR-25 | coverage_metrics | Metrics quantifying credible interval coverage | Dict[str, float] | Percentage | [0, 100] | None |
| ATTR-26 | bias_metrics | Metrics quantifying systematic bias in estimates | Dict[str, float] | Various | None | None |
| ATTR-27 | sample_size | Number of cells used in recovery | int | Count | > 0 | None |
| ATTR-28 | noise_level | Noise level used in data generation | float | Dimensionless | >= 0 | 0.0 |

@@

| Relationship ID | From Object | To Object | Type | Cardinality | Description |
|-----------------|-------------|-----------|------|-------------|-------------|
| REL-01 | AnnData | CountMatrix | Composition | 1:N | AnnData contains multiple count matrices (X, layers) |
| REL-02 | PyroVelocityModel | ModelParameters | Association | 1:1 | Model uses parameters for computation |
| REL-03 | PyroVelocityModel | PosteriorSamples | Association | 1:1 | Model generates posterior samples |
| REL-04 | PosteriorSamples | VelocityEstimates | Derivation | 1:1 | Velocity is computed from posterior samples |
| REL-05 | AnnData | VelocityEstimates | Association | 1:1 | Velocity estimates are stored in AnnData |
| REL-06 | PyroVelocityModel | ModelState | Composition | 1:1 | Model contains state |
| REL-07 | ModelState | ComponentConfig | Composition | 1:N | State contains component configurations |
| REL-08 | ModelParameters | SyntheticData | Generation | 1:N | Parameters are used to generate synthetic data |
| REL-09 | SyntheticData | ParameterRecoveryResults | Association | 1:1 | Synthetic data is used for parameter recovery validation |
| REL-10 | ParameterRecoveryResults | ModelParameters | Validation | N:1 | Recovery results validate parameter inference |

```{mermaid}
classDiagram
    AnnData *-- CountMatrix : contains
    PyroVelocityModel o-- ModelParameters : uses
    PyroVelocityModel --> PosteriorSamples : generates
    PosteriorSamples --> VelocityEstimates : computes
    AnnData o-- VelocityEstimates : stores
    PyroVelocityModel *-- ModelState : contains
    ModelState *-- ComponentConfig : contains
    ModelParameters --> SyntheticData : generates
    SyntheticData --> ParameterRecoveryResults : validates
    ParameterRecoveryResults --> ModelParameters : validates

@@

    class ComponentConfig {
        +str name
        +Dict params
    }

    class SyntheticData {
        +Tensor u_obs
        +Tensor s_obs
        +Dict true_parameters
        +int sample_size
        +float noise_level
    }

    class ParameterRecoveryResults {
        +Dict true_parameters
        +Dict estimated_parameters
        +Dict credible_intervals
        +Dict recovery_metrics
        +Dict coverage_metrics
    }
```
